GPROF-NN: A neural-network based implementation of the Goddard Profiling Algorithm

### 2021-05-28 article/manuscript.tex
Writing.

The GPROF-NN 0D algorithm is trained to predict all retrieval variable in
parallel by minimizing the sum of all losses. Each network is trained for
for 70 epochs with the Adam optimizer and a cosine annealing learning rate
schedule. Warm restarts are performed after 10, 20 and 40 epochs. The initial
learning rates for the first and second part of the training set to $5\cdot 10^{-4}$
and reduced to $10^{-4}$ for its last part.

learning rate is

@@

>>>>

The GPROF-NN 0D algorithm is trained to predict all retrieval variables in
parallel by simply minimizing the sum of all losses. Each network is trained for
for 70 epochs using the Adam optimizer and a cosine annealing learning rate
schedule. Warm restarts are performed after 10, 20 and 40 epochs. The initial
learning rates for the first and second part of the training set to $5\cdot
10^{-4}$ and reduced to $10^{-4}$ for its last part.

\subsubsection{Hyperparameter tuning}

To find a well-performing model, a range of configurations of the network
architecture described in Fig.~\ref{fig:gprof_nn_0d} has been trained on roughly
$10\ \unit{\%}$ of the training data and evaluated on a similar fraction of the
validation data. Each model configuration has been trained once as a QRNN and
once as a DRNN. The resulting mean percentage (MPE) error and mean absolute
percentage errors (MAPE) for a subset of the retrieval variables are displayed
in Fig.~\ref{fig:gprof_nn_0d_tuning}.

@@

The network architecture for the GPROF-NN 2D algorithm is a fully-convolutional
asymmetric encoder-decoder architecture followed by a separate network head for
each retrieval variable. Each stage of the encoder consists of a downsampling
layer followed by $N_B$ convolution blocks, each of which consists of two
exception layers followed by layer activation and GELU activation functions. The
input and outputs of each of those blocks are connected by a residual
connection. The decoder stages contain only a single of those blocks per layer.
Additional skip connections are inserted to provide a short cut between
respective stages of the encoder and decoder.

\subsubsection{Data augmentation}

Two properties make the application of a convolutional neural network to the GMI
observations peculiar: First, ground truth data is available only at the center
of the input images where the GMI observations overlap with DPR observations.
Secondly, the GMI observations are, due to the conical scanning of the sensor,
not invariant with respect to translations. This means that when the GMI
observations are interpreted as a rectilinear image any observed structure will
look different at the center of the swath compared to the edges. We found that
naive training of a convolutional neural network with this data can lead to
artifacts outside the swath center. To avoid these, the original images in the
training data are augmented on-the-fly with remapped versions that simulate the
effect of the relative placement of the observations in the swath. In addition
to this, observations from channels 10 and up are randomly set to missing in the
outer edges of the swath to simulate the effect of the missing observations that
effects the outer edges of the GMI swath.
